CE888 Assignment 2

### 2019-04-25 paper2/main.tex
final draft

%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{CE888 Assignment 1}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
The traditional unsupervised machine learning algorithms like k-Means, PCA and other
linear factor models are proven to be very useful for many important applications.
But in this day and age we are able to collect very complex, high dimensional data
and these simple but efficient algorithms are failing to learn or represent it\cite[p. 151]{goodfellow2016deep} In this
work we are going to use autoencoders to reduce the dimensions of three different datasets
and try to extract good features with and without softmax activation function. Then finally, 
we will cluster the data by using the extracted features and compare the results.

\end{abstract}

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{U}{nsupervised} learning is a branch of machine learning that deals with unlabeled data to find patterns, groups
and relations. This sub section of machine learning is useful to describe and understand unknown data, which 
also helps us understand the properties of future data samples. Right now we are collecting data much faster than we can process it.
And most of the time the data we acquire is unlabeled, and these huge amounts of data are complex, noisy
and it might include unnecessary information.\par There are different ways to reduce the complexity of a given data and learn good 
features from it. In this work, we will look at autoencoders, one of those procedures used to learn features from a given data.
Autoencoders are a part of representational learning, which can be trained with or without labels.
They play a key role in deep learning applications. They are used in dimensionality reduction and feature 
extraction, and together with deep learning they gained a lot of traction after Hinton proposed a method to train 
deep belief networks one layer at a time \cite{hinton2006fast}. Another addition to the autoencoder will be the softmax activation function.
Since softmax gives us the probability distribution over-the classes of our data- the output vector \cite[pp. 178-181]{goodfellow2016deep}, we can use the output of our
encoder for cluster assignment. The motivation for using autoencoders is, in what they do they are
very similar to traditional feature selection algorithms, but they are also capable of learning good features from
complex and high dimensional data. This aspect of autoencoders and unsupervised learning in general is important. 
As LeCun states in \cite{lecun2015deep}, people are not paying attention to unsupervised learning because of the success of the supervised learning
in recent years. But it will be far more important in the future because of how living beings perceive and learn, they do not
need labels to recognize the world around them.\par In this paper we are going to look at previous work (Section II),
look at different methods and present their results to show how they compare. In Section III, we are going to present 
our goals and what we want to accomplish. Then we will talk about the datasets that we are going to use for the task and 
give detailed information about them. The planned experiments and analyses for the task will be mentioned in Section IV.
Evaluation methods that are going to be used after the experiments will be presented in Section V.

\section{Background}
Clustering is an important part of data analysis, that is used to naturally group the samples in order to extract information or make decisions\cite{jain1999data}.
Autoencoders were discovered in 1985 and presented in\cite{rumelhart1985learning}. As the technology and our computational power improved immensely in the last decade, they became more and more feasible. After Hinton's solution\cite{hinton2006fast} these simple network models became very popular, now they are used in the state-of-the-art machine learning models, breaking records and in some tasks performing better than humans\cite{schmidhuber2015deep}. But most of the success and interest to the autoencoders can be attributed to the studies on supervised learning.\par
In this study, we will be looking at the other side and see how autoencoders perform compared to some of the well-known traditional models. These algorithms will be k-Means\cite{macqueen1967some} and DBSCAN\cite{ester1996density}.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=2.5in]{images/kM_DBSCAN.png}
  \caption{Clusters created by k-Means and DBSCAN.
  First Column : k-Means, Second Column : DBSCAN \newline
  Adapted from CE888 lab6 results.
  }
  \label{fig_cluster}
  \end{figure}

As it can be seen in Figure 1, k-Means algorithm cannot cluster non-linear data. Results in \cite{xie2016unsupervised} also confirm that, but it also shows us that the k-Means can be significantly improved if the dimensions of the data is reduced before the clustering procedure. In \cite{salakhutdinov2007learning} a multilayer network is used to map high dimensional, non-linear features to low-dimensional space which is fine-tuned to work with KNN. They achieved 1.01\% error rate using 7 nearest neighbors on MNIST dataset.\par

The study \cite{hinton2006reducing} gives a lot of insight on autoencoders and dimensionality reduction. They introduce the pre-training notion, which involves an effective way of initializing good weights and training the network layer by layer. Then they suggest fine-tuning the deep network as a whole. This practice can also be seen in \cite{vincent2010stacked} and \cite{xie2016unsupervised}, these studies also suggest pre-training. On the other hand, \cite{ma2015deep} concludes that the pre-training procedure does not improve the accuracy. Their experiments show that the pre-training, over multiple DNNs and 15 different datasets, on average reduce the accuracy. Another study\cite{poultney2007efficient} learns overcomplete representations of features for data reconstructions. It is shown in their results that this sparse, energy based autoencoder with linear layers can reproduce the MINST dataset with 0.7\% error, while not requiring any preprocessing.\cite{le2011building} also follows a similar approach, in which a deep sparse network trained in unsupervised fashion to recognize faces, humans and objects.
Finally, the last study we looked at implemented and compared different models (Figure 2). As a result \cite{coates2011analysis} states that the best result obtained came from k-Means clustering.

\begin{table}[!ht]
  %% increase table row spacing, adjust to taste
  \renewcommand{\arraystretch}{1.3}
  % if using array.sty, it might be a good idea to tweak the value of
  % \extrarowheight as needed to properly center the text within the cells
  \caption{Test recognition accuracy (and error) for
  NORB dataset.\newline
  Adapted from : \cite{coates2011analysis} }
  \label{tab_coates_res}
  \centering
  %% Some packages, such as MDW tools, offer better commands for making tables
  %% than the plain LaTeX2e tabular which is used here.
  \begin{tabular}{|c||c|}
  \hline
  Algorithm  &     Accuracy (error) \\
  \hline
  \hline
  Conv. Neural Network &    93.4\% (6.6\%)\\
  Deep Boltzmann Machine &      92.8\ (7.2\%)\\
  Deep Belief Network&    95.0\% (5.0\%)\\
  Deep neural network &    97.13\% (2.87\%)\\
  \hline
  Sparse auto-encoder &    96.9\% (3.\%)\\
  Sparse RBM &    96.2\ (3.8\%)\\
  K-means clustering ( 4000 features) &   97.21\% (2.79\%)\\
  \hline
  \end{tabular}
  \end{table}

\section{Methodology}
Our main goal in this study is to correctly cluster data using unsupervised machine learning techniques.
To accomplish this task, we will go through three different subtasks. First, we will use traditional clustering algorithms 
to see how they perform on our datasets. In the second part, we will use autoencoders to learn good features
from our datasets, then we will pass the- extracted features- outputs of the autoencoder to a clustering algorithm to see the difference
in their performance. Finally, we will change the output activation function of the autoencoder to the softmax 
activation function, which will give us the probability distribution of the output vector. Then we will use
the maximum weighted class as our cluster assignment.\par
Before starting the experiments, the datasets will be examined and processed appropriately for each sub-task.
For the first sub task, depending on the complexity and dimensions of the given data, a suitable clustering algorithm will be chosen.
For the second sub-task, features of the datasets will be standardized. As LeCun concludes in \cite{lecun2012efficient}, networks
can learn better and faster if the inputs are centered around zero with equalized covariance. Another consideration for sub-tasks
two and three is using ensemble of autoencoders. Since the optimization of the randomly initialized weights is 
difficult \cite{hinton2006reducing}, we can have multiple autoencoders in our model to mitigate the effects of random weight initialization.
For the last sub-task, since we will be using softmax as our activation function, the targets of our datasets will be 
converted to one hot coded versions.\par

@@

  \begin{table}[!ht]
  %% increase table row spacing, adjust to taste
  \renewcommand{\arraystretch}{1.3}
  % if using array.sty, it might be a good idea to tweak the value of
  % \extrarowheight as needed to properly center the text within the cells
  \caption{Dataset statistics.}
  \label{tab_datasets}
  \centering
  %% Some packages, such as MDW tools, offer better commands for making tables
  %% than the plain LaTeX2e tabular which is used here.
  \begin{tabular}{|c||c|c|c|}
  \hline
  Datasets &    Data Points &  Features &  Classes \\
  \hline
  \hline
  MNIST &    70.000 &   784 &                 10 \\
  \hline
  HAR &       10.299 &   561 &                 6 \\
  \hline
  HTRU2 &    17,898 &   8 &                  2 \\
  \hline
  \end{tabular}
  \end{table}
\section{Experiments}

In the experiments the implementation of the tasks and data analysis will be done using python. In the implementation of the traditional clustering algorithms, scikit-learn library will be used because of its easy to use nature and rich functions. For the rest of the tasks, TensorFlow and Keras libraries will be used to implement autoencoders, since TensorFlow converts the python code into C code in runtime, it is fast and it can run on GPUs, making it one of the best neural network library. Finally, the data analysis and plotting will be done using matplotlib and seaborn libraries.

\subsection{Clustering Algorithms}
In the core part of the clustering experiments, k-Means and DBSCAN algorithms will be used. These algorithms will be first compared to each other, then their results will be analyzed, evaluated and recorded for future comparisons. If there is enough time to deviate from the core path, agglomerative clustering algorithm will also be used, and these clustering algorithms will be tested again after applying dimensionality reduction methods like PCA/kernel PCA, LDA to the datasets. Given enough time, we will try these dimension reduction algorithms to confirm\cite{baldi1989neural}, in which it is concluded that the linear autoencoders with squared loss functions behave like PCAs. Otherwise, the experiments will continue with the autoencoders.

\subsection{AutoEncoder and Softmax}
Rest of the tasks share a similar model. Both of them will be using autoencoders to extract features. Implementation priority will be given to the main parts. First, an autoencoder will be trained and optimized to learn good features, then the output of these features will be passed to the mentioned clustering algorithms in the first task. Then, in the second part the output activations will be swapped with softmax activation function for clustering.\par
After implementing the minimum requirements, if there is enough time, various additional experiments will be performed. First, we will use linear autoencoder with one hidden layer for clustering to compare the performance to clustering with PCA. Then, we will perform the same experiment for non-linear autoencoder and kernel PCA. After the comparisons, we will try methods like Dropout\cite{srivastava2014dropout} and denoising, to prevent one to one mapping and promote generalization\cite{vincent2010stacked}. Finally, we will experiment with ensemble of autoencoders for more robust results.

\subsection{Similar Work}

In the study\cite{hinton2006reducing}, autoencoders are compared to non-linear PCA and other dimensionality reduction methods. And it is stated that the autoencoders outperform PCA, logistic PCA, latent semantic analysis (LSA) and local linear embedding techniques. The results of the reconstructions of MNIST digits from the study can be seen below.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=2.5in]{images/hinton_mnist.png}
  \caption{Top to bottom: Random test images, autoencoder reconstruction, logistic PCA reconstruction and standard PCA reconstruction.\newline Adapted from : \cite{hinton2006reducing}}
  \label{fig_hinton_mnist}
  \end{figure}

  \par

Another study \cite{xie2016unsupervised} uses autoencoders with dropout to extract good features for clustering.
In their model, softmax is applied to the resulting output vector from the autoencoder and then they optimize the network by comparing the soft assignment to the target clusters. Their results on MNIST can be seen below.

\begin{table}[!ht]
  %% increase table row spacing, adjust to taste
  \renewcommand{\arraystretch}{1.3}
  % if using array.sty, it might be a good idea to tweak the value of
  % \extrarowheight as needed to properly center the text within the cells
  \caption{Clustering accuracy on different subsamples of MNIST\newline
  Adapted from: \cite{xie2016unsupervised}}
  \label{tab_res_clus}
  \centering
  %% Some packages, such as MDW tools, offer better commands for making tables
  %% than the plain LaTeX2e tabular which is used here.
  \begin{tabular}{|c||c|c|c|c|c|c|}
  \hline
   Method  &    0.1 &  0.3 &  0.5 & 0.7 &  0.9  \\
  \hline
  \hline
  k-Means &    47.14\% &   49.93\% &     53.65\%   &    54.16\%  &   54.39\% \\
  \hline
  AE + k-Means & 66.82\% &   74.91\% &     77.93\%   &    80.04\%  &   81.31\% \\
  \hline
  DEC(proposed) &  70.10\% &   80.92\% &     82.68\%   &    84.69\%  &   85.41\% \\
  \hline
  \end{tabular}
  \end{table}

% Note that the IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command,
% and the \label for the overall figure must come after \caption.
% \hfil is used as a separator to get equal spacing.
% Watch out that the combined width of all the subfigures on a 
% line do not exceed the text width or a line break will occur.

% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat[]), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.
% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
% labels, the optional argument to \subfloat must be present. If a
% subcaption is not desired, just leave its contents blank,
% e.g., \subfloat[].

% An example of a floating table. Note that, for IEEE style tables, the
% \caption command should come BEFORE the table and, given that table
% captions serve much like titles, are usually capitalized except for words
% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
% and up, which are usually not capitalized unless they are the first or
% last word of the caption. Table text will default to \footnotesize as
% the IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
% \begin{table}[!t]
% %% increase table row spacing, adjust to taste
% \renewcommand{\arraystretch}{1.3}
% % if using array.sty, it might be a good idea to tweak the value of
% % \extrarowheight as needed to properly center the text within the cells
% \caption{An Example of a Table}
% \label{table_example}
% \centering
% %% Some packages, such as MDW tools, offer better commands for making tables
% %% than the plain LaTeX2e tabular which is used here.
% \begin{tabular}{|c||c|}
% \hline
% One & Two\\
% \hline
% Three & Four\\
% \hline
% \end{tabular}
% \end{table}

\section{Discussion}

At the end of the experiments, the evaluation of each model's cluster assignments will be calculated by using the V-measure. Since it takes into account both completeness and homogeneity of the clusters, it will be our choice of evaluation\cite{rosenberg2007v}. On the other hand we will also use the Fowlkes-Mallows score. Because it allows us to compare the performance of our three different methods\cite{fowlkes1983method}. Also, as an addition, we plan to evaluate the contingency matrix of each model, then perform McNemar's test between the models to see the significance of their difference.\par

This study will improve our understanding of autoencoders and it will give us a chance to observe how they perform compared to the traditional methods. It is assumed that the task of optimizing and fine tuning the encoders will be difficult and time consuming. Another benefit will be the comparison of the softmax clustering to the rest of the methods. Downside of this study will be understanding why the autoencoder does what it does, since it is a black box model. Luckily this is out of the scope of this study.

In a broader perspective, these experiments will improve our understanding of clustering, data manipulation and unsupervised learning in general. Understanding the data, its complexity and dimensions are not trivial yet very important. Even though Machine Learning and Data Science fields can provide many models and tools, it is difficult to get good results if we are not familiar with the data/problem we are dealing with.\par

>>>>

%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{CE888 Assignment 2}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
In this work, we explore autoencoders with a hidden softmax layer as a clustering method, by training an autoencoder and using the maximum of the output vector as our cluster assignment. 
We also examine traditioanl clustering algorithms and experiment with feature selection methods to see how they perform compared to each other. 
\end{abstract}

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{U}{nsupervised} learning is a branch of machine learning that deals with unlabeled data to find patterns, groups
and relations. This sub section of machine learning is useful to describe and understand unknown data, which 
also helps us understand the properties of future data samples. Right now we are collecting data much faster than we can process it.
And most of the time the data we acquire is unlabeled, and these huge amounts of data are complex, noisy
and it might include unnecessary information.\par There are different ways to reduce the complexity of a given data and learn good 
features from it. In this work, we looked at PCA and autoencoders. 
Autoencoders are a part of representational learning, which can be trained with or without labels.
They play a key role in deep learning applications. They are used in dimensionality reduction and feature 
extraction, and together with deep learning they gained a lot of traction after Hinton proposed a method to train 
deep belief networks one layer at a time \cite{hinton2006fast}. Another addition to the autoencoders and PCA was assigning the data to clusters usnig the softmax activation function.
Since softmax gives us the probability distribution over-the classes of our data- the output vector \cite[pp. 178-181]{goodfellow2016deep}, we can use the output of our
encoder for cluster assignment. The motivation for using autoencoders is, in what they do they are
very similar to traditional feature selection algorithms, but they are also capable of learning good features from
complex and high dimensional data. This aspect of autoencoders and unsupervised learning in general is important. 
As LeCun states in \cite{lecun2015deep}, people are not paying attention to unsupervised learning because of the success of the supervised learning
in recent years. But it will be far more important in the future because of how living beings perceive and learn, they do not
need labels to recognize the world around them.\par In this paper  In Section III, We explain the tools we used and mentioned the implementation steps we took for the experiments. Then we present the datasets that we used for the task and give detailed information about them.The detailed experiments and analyses for the task will be mentioned in Section IV.
Finally, we talk about how we evaluated our models and we discuss the results.

\section{Background}
Clustering is used to naturally group the samples in order to extract information or make decisions\cite{jain1999data}.
Autoencoders were discovered in 1985 and presented in\cite{rumelhart1985learning}. As the technology and our computational power improved immensely in the last decade, they became more and more feasible. After Hinton's solution\cite{hinton2006fast} these simple network models became very popular, now they are used in the state-of-the-art machine learning models, breaking records and in some tasks performing better than humans\cite{schmidhuber2015deep}. But most of the success and interest to the autoencoders can be attributed to the studies on supervised learning.\par
In this study, first we looked at the well-known traditional models k-Means\cite{macqueen1967some} and DBSCAN\cite{ester1996density}. K-Means tries to divide the data into k clusters which are defined by k centroids. Each sample given to the algorithm is assigned to the cluster with the nearest mean. The k-Means algorithm is not deterministic and results in different clusters for different initializations. DBSCAN on the other hand separates the data by looking at it as highly populated areas that are separated by low population areas. The DBSCAN algorithm, as opposed to k-Means, is deterministic and results in same clusters.

After the traditional methods, we trained different autoencoders for dimensionality reduction and clustering. The model of our autoencoder networks is a simple end to end auto encoder with one hidden layer.
\begin{figure}[!ht]
  \centering
  \includegraphics[width=2.5in]{images/ae_model.png}
  \caption{Autoencoder model
  }
  \label{fig_ae_model}
  \end{figure}
First model uses linear activation functions, in order to achieve PCA-like behavior. Then our second autoencoder model uses non-linear sigmoid activation function, for the purpose of learning good features from high-dimensional data. 
\begin{equation}
  \sigma(a)=\frac{1}{1+e^{-a}}
\end{equation}
The output of these two models are used as the inputs to the traditional clustering algorithms. 
Finally, our last model uses a softmax activation function in the latent feature layer. Softmax provides the probability distribution over the nodes of our latent feature layer. Which are then used to cluster the data
\begin{equation}
  softmax_i(a)=\frac{\exp{a_i}}{\sum\exp{a_i}}
\end{equation}

Even though the k-Means algorithm cannot cluster non-linear data, \cite{xie2016unsupervised} and \cite{salakhutdinov2007learning} show that traditional clustering methods can be improved significantly if a good feature selection method is applied before the clustering procedure. In \cite{hinton2006reducing} feature selection with autoencoders are compared to other feature selection methods like PCA , non-linear PCA etc. And it is stated that the autoencoders perform better than these techniques if they are initialized and trained carefully. Another study\cite{poultney2007efficient} learns overcomplete representations of features for data reconstructions. It is shown in their results that this sparse, energy based autoencoder with linear layers can reproduce the MINST dataset with 0.7\% error, while not requiring any preprocessing.\cite{le2011building} also follows a similar approach, in which a deep sparse network trained in unsupervised fashion to recognize faces, humans and objects..
  \par
The most similar work \cite{xie2016unsupervised} uses stacked denoising autoencoders to extract good features for clustering. When the pre-training of their SAE is over, the trained encoder model is extracted and merged with a custom layer which applies softmax activation to the resulting output vector from the encoder. After setting up the final model, they use k-Means to generate cluster centroids for the data. Then, these centroids are used to train/fine-tune the final model by minimizing the KL-divergence between the soft assignment of the network and the cluster centroids created by k-Means.

\section{Methodology}
In this study, we experimented with autoencoders using a softmax latent feature layer, to bin the data into the softmax nodes and use the maximum of this vector for our clustering assignment. But before testing our main objective, we first set up a base line specific for our datasets using the traditional clustering algorithms. We also looked into how feature selection procedures affect the performance of these clustering algorithms. First with PCA, then with linear autoencoders and finally with non-linear autoencoders. After creating the baseline and experimenting with feature selection, we proceeded to our main objective.
\par
Implementation of the tasks were done using Python. Reading the data files and importing procedures were implemented using the Pandas library. The traditional clustering algorithms were implemented using the scikit-learn library which already includes the k-Means and DBSCAN algorithms. Also some data processing operations like normalization/scaling, train-test split were done using the functions provided by scikit-learn. For the tasks related to autoencoders, TensorFlow and Keras libraries were used. Then, a high level autoencoder class was implemented in order to easily create autoencoders with desired layers, nodes and activation functions. Finally, the data presentation and plotting the results were done using matplotlib and seaborn libraries.
\par

@@

  \begin{table}[!ht]
  %% increase table row spacing, adjust to taste
  \renewcommand{\arraystretch}{1.3}
  % if using array.sty, it might be a good idea to tweak the value of
  % \extrarowheight as needed to properly center the text within the cells
  \caption{Dataset statistics.}
  \label{tab_datasets}
  \centering
  %% Some packages, such as MDW tools, offer better commands for making tables
  %% than the plain LaTeX2e tabular which is used here.
  \begin{tabular}{|c||c|c|c|}
  \hline
  Datasets &    Data Points &  Features &  Classes \\
  \hline
  \hline
  MNIST &    70.000 &   784 &                 10 \\
  \hline
  HAR &       10.299 &   561 &                 6 \\
  \hline
  HTRU2 &    17,898 &   8 &                  2 \\
  \hline
  \end{tabular}
  \end{table}
\section{Experiments}
The experiments done in this work can be separated into three sub-groups. First part is the clustering algorithms. The second part is the effect of feature selection methods. And the third part is the clustering using the maximum of the softmax output vector. Before running the experiments, for repeatability purposes, the randomness states were fixed.

\subsection{Clustering }

First, to have a base-line to compare our results we clustered our data using the k-Means algorithm. Since our datasets are labeled, we set the number of centroids k to the number of classes for each dataset, and we set the number of initializations to 10.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=3.3in]{images/pulsarData_Cluster.png}
  \caption{Plot of first two features of the HTRU dataset, clustered and labeled by k-Means}
  \label{fig_pulsarData_Cluster}
\end{figure}
\par

\begin{table}[!ht]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{K-Means Clustering Results.}
\label{tab_kmeans_res}
\centering
\begin{tabular}{|c|c|c|c|c|}
\hline
Data - Method &  Homogenity &  Complete. &  V-Measure &  Accuracy \\
\hline
\hline
htru k-Means &    0.031206 &      0.023027 &   0.026500 &  0.234949 \\
\hline
har k-Means&    0.570123 &      0.600019 &   0.584689 &  0.289853 \\
\hline 
mnist k-Means&    0.486553 &      0.497699 &   0.492063 &  0.083400 \\
\hline
\end{tabular}
\end{table}

\begin{figure}[!ht]
  \centering
  \includegraphics[width=3.3in]{images/harData_Cluster.png}
  \caption{Plot of first two features of the HAR dataset, clustered and labeled by k-Means}
  \label{fig_harData_Cluster}
\end{figure}
\par

In addition to the k-Means, DBSCAN algorithm was also used to cluster our datasets. But the DBSCAN algorithm provided by the scikit-learn library labeled our datasets as noise and failed to cluster them. Which we think is because of the high dimensionality of our data. Thus, the results of the DBSCAN algorithm were discarded. Also because of the poor results of the MNIST, all the centroids were stacked at the origin and its k-Means plot was not included here.

\begin{figure*}[!htb]
  \minipage{0.32\textwidth}
    \includegraphics[width=\linewidth]{images/pulsarpca.png}
    \caption{HTRU Explained variance ratio.}\label{fig_pulsarpca}
  \endminipage\hfill
  \minipage{0.32\textwidth}
    \includegraphics[width=\linewidth]{images/harpca.png}
    \caption{HAR Explained variance ratio.}\label{fig_harpca}
  \endminipage\hfill
  \minipage{0.32\textwidth}%
    \includegraphics[width=\linewidth]{images/mnistpca.png}
    \caption{MNIST Explained variance ratio.}\label{fig_mnistpca}
  \endminipage
  \end{figure*}

\begin{figure*}[!htb]
  \minipage{0.32\textwidth}
    \includegraphics[width=\linewidth]{images/pca_pulsarData_Cluster.png}
    \caption{HTRU clustering after PCA.}\label{fig_pca_pulsarData_Cluster}
  \endminipage\hfill
  \minipage{0.32\textwidth}
    \includegraphics[width=\linewidth]{images/pca_harData_Cluster.png}
    \caption{HAR clustering after PCA}\label{fig_pca_harData_Cluster}
  \endminipage\hfill
  \minipage{0.32\textwidth}%
    \includegraphics[width=\linewidth]{images/pca_mnistData_Cluster.png}
    \caption{MNIST clustering after PCA}\label{fig_pca_mnistData_Cluster}
  \endminipage
  \end{figure*}
  \begin{table*}[!ht]
    %% increase table row spacing, adjust to taste
    \renewcommand{\arraystretch}{1.3}
    % if using array.sty, it might be a good idea to tweak the value of
    % \extrarowheight as needed to properly center the text within the cells
    \caption{K-Means Clustering Results After PCA.}
    \label{tab_kmeans_pca_res}
    \centering
    \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    Data - Method & Num. Components & Homogenity &  Completeness &  V-Measure &  Accuracy \\
    \hline
    \hline
    htru k-Means w/PCA &  4  &0.031206 &  2.302700e-02 &  2.649971e-02 &  0.765051 \\
    \hline
    har k-Means w/PCA&  200  & 0.570146 &  5.996766e-01 &  5.845388e-01 &  0.318961 \\
    \hline 
    mnist k-Means w/PCA& 400 & 0.486113 &  4.971782e-01 &  4.915831e-01 &  0.072782 \\
    \hline
    \end{tabular}
    \end{table*}

%In the core part of the clustering experiments, k-Means and DBSCAN algorithms will be used. These algorithms will be first compared to each other, then their results will be analyzed, evaluated and recorded for future comparisons. If there is enough time to deviate from the core path, agglomerative clustering algorithm will also be used, and these clustering algorithms will be tested again after applying dimensionality reduction methods like PCA/kernel PCA, LDA to the datasets. Given enough time, we will try these dimension reduction algorithms to confirm\cite{baldi1989neural}, in which it is concluded that the linear autoencoders with squared loss functions behave like PCAs. Otherwise, the experiments will continue with the autoencoders.

\subsection{Feature Selection and Clustering}
\subsubsection{PCA}
For the feature selection, the first method we implemented was PCA. We transformed our data using the PCA function and retrieved the explained variance ratio to decide how many components we need. As it can be seen in figures 7,8 and 9, we choose the least amount of components representing almost all of the data.($\cong$0.99\%) As a result, we used 4 components for HTRU, 200 for HAR and 400 for MNIST and clustered the PCA reduced data using the k-Means algorithm. 
\par
\subsubsection{Linear AE}
After clustering with PCA, we implemented an end-to-end autoencoder with linear activation functions. The autoencoder has the basic structure shown in Figure 1. The input and output layers were set to the number of features of the datasets, and the latent feature layers were set to the same numbers we used in the PCA reduction. 'Adam' was used as the optimizer, and 'mean squared error' was used as the loss function. Finally, the batch size was set as 128, number of training epochs were set to 50 and our test sets were assigned as the validation data. \par
\subsubsection{Non-linear AE}
For the last feature selection method, we implemented the same autoencoder but replaced the linear activation function in the encoder model with a non-linear activation function, which is in this case was sigmoid. We also increased the number of training epochs for the non-linear case to 100. Previous number of training epochs was not enough for the non-linear autoencoder, which ended up as an underfitted model.

\begin{figure*}[!htb]
  \minipage{0.32\textwidth}
    \includegraphics[width=\linewidth]{images/aelinear_pulsarData_Cluster.png}
    \caption{HTRU Explained variance ratio.}\label{fig_aelinear_pulsarData_Cluster}
  \endminipage\hfill
  \minipage{0.32\textwidth}
    \includegraphics[width=\linewidth]{images/aelinear_harData_Cluster.png}
    \caption{HAR Explained variance ratio.}\label{fig_aelinear_harData_Cluster}
  \endminipage\hfill
  \minipage{0.32\textwidth}%
    \includegraphics[width=\linewidth]{images/aelinear_mnistData_Cluster.png}
    \caption{MNIST Explained variance ratio.}\label{fig_aelinear_mnistData_Cluster}
  \endminipage
  \end{figure*}

\begin{figure*}[!htb]
  \minipage{0.32\textwidth}
    \includegraphics[width=\linewidth]{images/aenon-linear_pulsarK_Means_Data_Cluster.png}
    \caption{HTRU clustering after PCA.}\label{fig_linear_pulsarK_Means_Data_Cluster}
  \endminipage\hfill
  \minipage{0.32\textwidth}
    \includegraphics[width=\linewidth]{images/aenon-linear_harK_Means_Data_Cluster.png}
    \caption{HAR clustering after PCA}\label{fig_aenon-linear_harK_Means_Data_Cluster}
  \endminipage\hfill
  \minipage{0.32\textwidth}%
    \includegraphics[width=\linewidth]{images/aenon-linear_mnistK_Means_Data_Cluster.png}
    \caption{MNIST clustering after PCA}\label{fig_aenon-linear_mnistK_Means_Data_Cluster}
  \endminipage
  \end{figure*}
  \begin{table*}[!ht]
    %% increase table row spacing, adjust to taste
    \renewcommand{\arraystretch}{1.3}
    % if using array.sty, it might be a good idea to tweak the value of
    % \extrarowheight as needed to properly center the text within the cells
    \caption{K-Means Clustering Results for linear and non-linear autoencoders.}
    \label{tab_lae_nlae}
    \centering
    \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    Data - Method & Num. Latent Features & Homogenity &  Completeness &  V-Measure &  Accuracy \\
    \hline
    \hline
    htru k-Means w/LinAE &  4  &0.421208 &  3.674883e-01 &  3.925187e-01 &  0.935326 \\
    \hline
    har k-Means w/LinAE&  200  & 0.548367 &  5.951099e-01 &  5.707828e-01 &  0.180495 \\
    \hline 
    mnist k-Means w/LinAE& 400 & 0.247627 &  2.516646e-01 &  2.496297e-01 &  0.087673 \\
    \hline
    \hline
    htru k-Means w/N-LinAE &  4  &0.361558 &      0.312014 &   0.334964 &  0.925059 \\
    \hline
    har k-Means w/N-LinAE&  200  & 0.542958 &      0.574462 &   0.558266 &  0.184576 \\
    \hline 
    mnist k-Means w/N-LinAE& 400 & 0.493461 &      0.500976 &   0.497190 &  0.148200 \\
    \hline
    \end{tabular}
    \end{table*}
    \subsection{Softmax Clustering}
    In our final experiment, since our objective is to 'bin' our data into the softmax nodes; We created a new auto encoder with a softmax latent feature layer that has N nodes where N is the number of classes the given dataset has. Then for the output layer we followed \cite{xie2016unsupervised} and used Rectifier Linear Unit(ReLU) activation function, and we kept the rest of the network parameters same. After training the models, we separated each test set according to their class labels. Then we predicted each 'class-set' one by one to see which node played an active role in the clustering/binning of a given class sample. Finally, as it can be seen in the Table 5, if the classes of a given set was binned properly, we made the assumption that the model was trained correctly and evaluated the scores on all of the test set. The results presented here only show the nodes that best cluster a given class-set, excluding the HTRU set because it only has two classes. Complete class versus node accuracy tables can be found in the Appendix.

\begin{table}[!ht]
  %% increase table row spacing, adjust to taste
  \renewcommand{\arraystretch}{1.3}
  % if using array.sty, it might be a good idea to tweak the value of
  % \extrarowheight as needed to properly center the text within the cells
  \caption{HTRU Softmax Results.}
  \label{tab_htru_sfx}
  \centering
  \begin{tabular}{|c|c|c|}
  \hline
  Class vs. Output Node Accuracy& Node 0 & Node 1  \\
  \hline
  \hline
  Class 0   &  \textbf{0.9963}    & 0.00368 \\
  \hline
  Class 1   &  0.26168  &  \textbf{0.73831} \\
  \hline
  \multicolumn{2}{|c|}{HTRU Test Accuracy:  } & 0.973184\\
  \hline

\end{tabular}
  \end{table}
